### 2015-05-12 background.tex
update in background

Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput). In device drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core. Likewise, performance is also important: e.g. network drivers should be able to process incoming packages in a timely manner.

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\caption{Data races in a Linux driver that we discovered using \whoop}\label{fig:data_race_example}
\end{figure}

@@

We believe that our focus in locks is justified, because the majority of Linux drivers still use locks. As a sanity check, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations. We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\caption{Using a lock to eliminate the data races}\label{fig:lock_example}
\end{figure}

\subsection{Lockset analysis}
\label{bg:lockset}

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{TODO} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.

>>>>

Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput). In drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core. Likewise, performance is also important: e.g. network drivers should be able to process incoming packages in a timely manner.

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\caption{Example of data races in the generic\_nvram Linux char driver}\label{fig:data_race_example}
\end{figure}

@@

We believe that our focus in locks is justified, because the majority of Linux drivers still use locks. To check this, we used a simple grep command and found that 68\% of the drivers from the 4.0 Linux kernel distribution use locks, versus 13\% that use atomic operations. We also found that 31\% of the drivers use neither locking nor atomic primitives, which suggests that they are either simple (non-concurrent) drivers or use some other lock-free synchronization primitive.

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\caption{Using a lock to eliminate the data races in the previous example}\label{fig:lock_example}
\end{figure}

\subsection{Lockset Analysis}
\label{bg:lockset}

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.
